---
title: "The U.S. Drug Overdose Crisis"
author: "Your Name"
---

# The U.S. Drug Overdose Crisis

The United States is an outlier among high-income nations in one striking respect: unintentional injuries rank as the third leading cause of death, driven overwhelmingly by drug overdoses [@cdc_faststats]. In the most recent finalized data, unintentional injuries accounted for over 222,000 deaths annually [@cdc_faststats]. The primary driver is synthetic opioids, particularly illicitly manufactured fentanyl, which has transformed what began as a prescription opioid problem into a mass-casualty poisoning event.

No peer nation experiences anything comparable. The [CDC National Center for Health Statistics](https://www.cdc.gov/nchs/fastats/leading-causes-of-death.htm) ranks unintentional injuries (driven by overdoses) as the third leading cause of death in the United States, behind heart disease and cancer [@cdc_faststats]. The [Commonwealth Fund 2025 Scorecard on State Health System Performance](https://www.commonwealthfund.org/publications/scorecard/2025/jun/2025-scorecard-state-health-system-performance) highlights that premature, avoidable deaths are increasing in the United States while declining in comparable high-income countries, with drug overdose a major contributor to this divergence [@commonwealth2025].

This chapter explores what the data shows, where the crisis hits hardest, and how to access and analyze the underlying datasets and research literature programmatically.

## The Scale of the Problem

Drug overdose deaths in the United States have followed a steep, accelerating trajectory over the past two decades. The crisis is commonly described in three waves:

1. **Wave 1 (1990s)**: Prescription opioid deaths rise, driven by aggressive marketing of OxyContin and similar drugs
2. **Wave 2 (2010s)**: Heroin deaths surge as prescription access tightens and users shift to cheaper alternatives
3. **Wave 3 (2013 onward)**: Synthetic opioids, primarily illicitly manufactured fentanyl, overwhelm the death toll

The third wave is the dominant driver today. Fentanyl is 50 to 100 times more potent than morphine, is inexpensive to manufacture, and is increasingly found mixed into non-opioid drug supplies (cocaine, methamphetamine, counterfeit pills), catching users unaware.

## State-Level Variation

The overdose crisis does not hit uniformly. State-level overdose death rates in 2023 reveal dramatic geographic variation:

| State | Drug Overdose Death Rate (per 100,000) | Context |
|:---|---:|:---|
| New Mexico | 48.9 | Highest in the Four Corners region |
| Colorado | 30.6 | Acute crisis among people experiencing homelessness |
| Utah | 21.4 | Lower rate, but rising |
| National average | ~33 | Approximate |

New Mexico's rate (48.9 per 100,000) is among the highest in the nation and more than double Utah's rate, despite the states sharing a border. These disparities reflect differences in drug supply routes, poverty concentration, rural healthcare access, and harm reduction infrastructure. [America's Health Rankings](https://www.americashealthrankings.org/) provides additional state-level context: Utah ranks among the healthiest states overall but faces an anomalous suicide crisis, while Colorado's [2025 Death Review](https://www.coloradocoalition.org/sites/default/files/2025-12/2025%20Death%20Review_Final.pdf) by the Colorado Coalition for the Homeless found that drug overdoses account for 77% of deaths among people experiencing homelessness in Denver.

The [Colorado Consortium for Prescription Drug Abuse Prevention](https://corxconsortium.org/data/) maintains detailed data on the state's overdose landscape, including toxicological detail from the State Unintentional Drug Overdose Reporting System (SUDORS).

## Data Sources

The primary datasets for analyzing drug overdose mortality in the United States are:

### CDC WONDER

[CDC WONDER](https://wonder.cdc.gov/) (Wide-ranging ONline Data for Epidemiologic Research) is the premier federal system for querying mortality data from the National Vital Statistics System. It provides:

- Mortality counts, crude rates, and age-adjusted rates
- Stratification by geography (national, state, county), demographics, urbanization, and ICD-10 codes
- Multiple Cause of Death files that capture all substances listed on the death certificate

::: {.callout-warning title="Data Suppression"}
CDC WONDER suppresses counts below 10 deaths in any given cell to prevent identification. Analyses of small populations or rare drug categories at the county level will encounter this limitation.
:::

### CDC Open Data (WONDER API and Socrata)

CDC publishes several overdose-specific datasets on its open data portal at [data.cdc.gov](https://data.cdc.gov/), including provisional drug overdose death counts updated monthly. These datasets are Socrata-hosted and accessible via API.

### State-Level Portals

- **Colorado**: [CDPHE Open Data Portal](https://data.colorado.gov/) (Socrata), plus the State Unintentional Drug Overdose Reporting System (SUDORS) for toxicological detail
- **New Mexico**: [NM-IBIS](https://ibis.doh.nm.gov/) query system for death certificate data and emergency department visits
- **Utah**: [Utah Open Data Catalog](https://opendata.utah.gov/) (Socrata)

## Accessing the Data with R

### VSRR Provisional Drug Overdose Death Counts

The primary dataset used in this chapter is the [VSRR Provisional Drug Overdose Death Counts](https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a) (dataset ID: `xkb8-kh2a`), published by the [National Center for Health Statistics (NCHS)](https://www.cdc.gov/nchs/nvss/vsrr/drug-overdose-data.htm) as part of the Vital Statistics Rapid Release (VSRR) program. This dataset presents provisional counts of drug overdose deaths based on the current flow of mortality data in the [National Vital Statistics System (NVSS)](https://www.cdc.gov/nchs/nvss/index.htm). It is updated monthly, with data available within four months after the date of death.

Provisional counts may not include all deaths that occurred during a given period because drug overdose deaths often require lengthy investigation, including toxicology testing. Death certificates may initially be filed with manner of death "pending investigation" or with a preliminary cause of death. To address this undercount, NCHS developed statistical methods to generate **predicted provisional counts** that adjust for reporting delays based on historical completeness patterns. Drug overdose deaths are identified using ICD-10 underlying cause-of-death codes: X40--X44 (unintentional), X60--X64 (suicide), X85 (homicide), and Y10--Y14 (undetermined intent).

The dataset reports **12-month ending counts**, meaning each row represents the total number of deaths occurring in the 12-month period ending in the month indicated. For example, a row for June 2024 captures deaths from July 2023 through June 2024. This rolling window smooths out seasonal variation and provides a stable trend measure. The `indicator` field specifies which outcome is being counted: total drug overdose deaths, or deaths involving a specific drug or drug class, identified by ICD-10 multiple cause-of-death codes:

- **Heroin** (T40.1)
- **Natural opioid analgesics** including morphine, codeine, oxycodone, hydrocodone (T40.2)
- **Methadone** (T40.3)
- **Synthetic opioids other than methadone**, including fentanyl and tramadol (T40.4)
- **Cocaine** (T40.5)
- **Psychostimulants with abuse potential**, including methamphetamine (T43.6)

Because a single death may involve multiple substances, it can appear in more than one drug category. For example, a death involving both heroin and fentanyl would be counted under both heroin and synthetic opioids other than methadone.

#### Dataset Columns

The dataset contains 12 fields:

| Column | API Field | Type | Description |
|:---|:---|:---|:---|
| State | `state` | Text | Two-letter jurisdiction abbreviation (e.g., "US", "CO", "NM") |
| Year | `year` | Number | Calendar year of the 12-month ending period |
| Month | `month` | Text | Month name of the 12-month ending period (e.g., "January") |
| Period | `period` | Text | Label describing the time span (e.g., "12 Month-ending Provisional Counts") |
| Indicator | `indicator` | Text | The specific outcome measured: "Number of Drug Overdose Deaths" or a drug-specific category such as "Heroin (T40.1)" or "Synthetic opioids, excl. methadone (T40.4)" |
| Data Value | `data_value` | Number | The reported provisional count of deaths for the indicator, jurisdiction, and period |
| Predicted Value | `predicted_value` | Number | The model-adjusted count that corrects for reporting delays and pending investigations |
| Percent Complete | `percent_complete` | Text | Percentage of expected death records received and processed for the jurisdiction and period |
| Percent Pending Investigation | `percent_pending_investigation` | Number | Percentage of death records where manner of death is listed as "pending investigation" |
| State Name | `state_name` | Text | Full jurisdiction name (e.g., "United States", "Colorado") |
| Footnote | `footnote` | Text | Contextual notes about data quality or suppression for the observation |
| Footnote Symbol | `footnote_symbol` | Text | Symbol corresponding to the footnote (if any) |

::: {.callout-warning title="Interpreting Provisional Data"}
Provisional counts are often lower than final counts due to reporting delays. The `predicted_value` column provides a corrected estimate but may not fully account for unusually long delays in some jurisdictions. Comparisons of drug-specific death counts across states should be made cautiously because the percentage of death certificates with specific drugs recorded varies by jurisdiction. For definitive counts, use the [final mortality data](https://www.cdc.gov/nchs/nvss/mortality_public_use_data.htm) published annually through CDC WONDER.
:::

The following code demonstrates how to pull this dataset from CDC's Socrata-hosted open data portal.

```{r}
#| label: cdc-overdose-data
#| eval: true
#| code-fold: true

library(httr2)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(plotly)

# CDC provisional drug overdose death counts (Socrata endpoint)
# Dataset: VSRR Provisional Drug Overdose Death Counts
# Source: https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a

cdc_url <- "https://data.cdc.gov/resource/xkb8-kh2a.json"

# Query: US + all states, total drug overdose deaths only
# The Socrata API caps at 50,000 rows per request
response <- request(cdc_url) |>
  req_url_query(
    `$where`  = "indicator = 'Number of Drug Overdose Deaths'",
    `$limit`  = 50000,
    `$order`  = "year ASC, month ASC"
  ) |>
  req_perform()

overdose_raw <- response |>
  resp_body_json(simplifyVector = TRUE)
```

The query returned **`r nrow(overdose_raw)`** rows covering **`r n_distinct(overdose_raw$state_name)`** jurisdictions.

::: {.callout-tip title="Reproducibility"}
The Socrata API does not require authentication for public datasets. The dataset identifier `xkb8-kh2a` is stable. You can also download the full dataset as CSV from the [dataset page](https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a). For full methodology, including the statistical models used to generate predicted counts and the data quality thresholds applied to each jurisdiction, see the [Technical Notes](https://www.cdc.gov/nchs/nvss/vsrr/drug-overdose-data.htm).
:::

### Interactive Time Series: Drug Overdose Deaths by Jurisdiction

The plot below shows 12-month ending provisional drug overdose death counts for the United States and every reporting state. Use the legend to toggle individual jurisdictions on or off, hover over data points for exact counts, and use the range slider to zoom into specific time periods.

```{r}
#| label: overdose-time-series
#| eval: true
#| code-fold: true

# Month lookup for proper date ordering
month_lookup <- c(
  "January" = 1, "February" = 2, "March" = 3, "April" = 4,
  "May" = 5, "June" = 6, "July" = 7, "August" = 8,
  "September" = 9, "October" = 10, "November" = 11, "December" = 12
)

overdose_ts <- overdose_raw |>
  mutate(
    data_value     = as.numeric(data_value),
    predicted_value = as.numeric(predicted_value),
    year_num       = as.integer(year),
    month_num      = month_lookup[month],
    date           = as.Date(paste(year_num, month_num, "01", sep = "-"))
  ) |>
  filter(!is.na(data_value)) |>
  arrange(state_name, date)

# Separate United States (always visible) from states (hidden by default)
us_data    <- overdose_ts |> filter(state_name == "United States")
state_data <- overdose_ts |> filter(state_name != "United States")

# Get sorted list of state names
state_names <- sort(unique(state_data$state_name))

# Build the plotly figure
p <- plot_ly() |>
  # United States trend line (bold, always visible)
  add_trace(
    data = us_data,
    x = ~date, y = ~data_value,
    type = "scatter", mode = "lines",
    name = "United States",
    line = list(color = "#2494f7", width = 3),
    hovertemplate = paste(
      "<b>United States</b><br>",
      "Period ending: %{x|%B %Y}<br>",
      "Reported deaths: %{y:,.0f}<extra></extra>"
    )
  )

# Add each state as a separate trace (hidden by default via legendgroup)
for (st in state_names) {
  st_df <- state_data |> filter(state_name == st)
  p <- p |>
    add_trace(
      data = st_df,
      x = ~date, y = ~data_value,
      type = "scatter", mode = "lines",
      name = st,
      visible = "legendonly",
      hovertemplate = paste0(
        "<b>", st, "</b><br>",
        "Period ending: %{x|%B %Y}<br>",
        "Reported deaths: %{y:,.0f}<extra></extra>"
      )
    )
}

p <- p |>
  layout(
    title = list(
      text = "12-Month Ending Provisional Drug Overdose Deaths by Jurisdiction",
      font = list(size = 16)
    ),
    xaxis = list(
      title = "12-Month Period Ending",
      rangeslider = list(visible = TRUE),
      type = "date"
    ),
    yaxis = list(
      title = "Deaths (12-month ending count)",
      separatethousands = TRUE
    ),
    legend = list(
      orientation = "v",
      x = 1.02, y = 1,
      font = list(size = 10)
    ),
    hovermode = "x unified",
    margin = list(r = 150)
  ) |>
  config(displayModeBar = TRUE)

p
```

## State Comparisons: Four Corners

The Four Corners states (Arizona, Colorado, New Mexico, Utah) present a natural comparison group: geographically contiguous, but with very different overdose profiles. We can use the same CDC dataset to compare state-level trends.

```{r}
#| label: four-corners-comparison
#| eval: false
#| code-fold: false

# Filter for Four Corners states
four_corners <- c("Arizona", "Colorado", "New Mexico", "Utah")

fc_response <- request(cdc_url) |>
  req_url_query(
    `$where`  = paste0(
      "state_name in('", paste(four_corners, collapse = "','"), "')",
      " AND indicator = 'Number of Drug Overdose Deaths'"
    ),
    `$limit`  = 5000,
    `$order`  = "year DESC, month DESC"
  ) |>
  req_perform()

fc_data <- fc_response |>
  resp_body_json(simplifyVector = TRUE) |>
  mutate(
    data_value = as.numeric(data_value),
    period = paste(year, month, sep = "-")
  )

# Plot: overdose death trends by state
ggplot(fc_data, aes(x = period, y = data_value, color = state_name, group = state_name)) +
  geom_line() +
  labs(
    title = "Drug Overdose Deaths: Four Corners States",
    x = "Period",
    y = "Deaths (12-month rolling count)",
    color = "State"
  ) +
  theme_minimal()
```

## What the Interventions Look Like

Public health responses to the overdose crisis operate at multiple levels:

**Harm reduction**: Expanding naloxone distribution (the opioid reversal drug), establishing syringe services programs, and drug checking services [@cdc_overdose_prevention]. Colorado has enacted harm reduction legislation expanding naloxone access via standing orders through pharmacies [@colorado_harm_reduction]. New Mexico operates state-funded harm reduction programs focused on overdose death reduction and participant education [@nmdoh_harm_reduction].

**Treatment access**: Increasing access to medications for opioid use disorder (MOUD), including buprenorphine, methadone, and naltrexone [@cdc_overdose_prevention]. Federal funding through State Opioid Response (SOR) grants ($1.48 billion nationally) and Tribal Opioid Response grants (~$63 million) support these efforts [@hhs_sor_grants].

**Surveillance**: Enhanced death investigation through SUDORS (State Unintentional Drug Overdose Reporting System) provides toxicological detail beyond what death certificates capture, revealing which specific substances are involved [@cdc_sudors].

## Federal Funding

Opioid response funding flows through several federal mechanisms. The [USAspending API](https://api.usaspending.gov/) enables programmatic tracking of these allocations:

```{r}
#| label: usaspending-opioid
#| eval: true
#| code-fold: true

# Query USAspending for opioid-related grants
# Documentation: https://api.usaspending.gov/

usa_spending_url <- "https://api.usaspending.gov/api/v2/search/spending_by_award/"

# Dynamic date window: past 5 years from today
date_end   <- Sys.Date()
date_start <- as.Date(paste0(as.integer(format(date_end, "%Y")) - 5, "-01-01"))
year_end   <- as.integer(format(date_end, "%Y"))
year_start <- year_end - 5

# POST request body: opioid-related keyword search for grants
request_body <- list(
  filters = list(
    keywords = list("opioid", "overdose", "substance abuse"),
    time_period = list(
      list(start_date = format(date_start, "%Y-%m-%d"),
           end_date   = format(date_end, "%Y-%m-%d"))
    ),
    award_type_codes = c("02", "03", "04", "05")
  ),
  fields = c(
    "Award ID", "Recipient Name", "Award Amount",
    "Awarding Agency", "Awarding Sub Agency",
    "Start Date", "End Date", "Description",
    "generated_internal_id"
  ),
  limit = 100,
  page = 1
)

# Paginate to collect up to 500 awards for word cloud analysis
# USAspending API caps at 100 results per page
all_results <- list()
for (pg in 1:5) {
  request_body$page <- pg
  spending_response <- request(usa_spending_url) |>
    req_body_json(request_body) |>
    req_perform()

  page_data <- spending_response |>
    resp_body_json(simplifyVector = TRUE)

  if (is.null(page_data$results) || nrow(page_data$results) == 0) break
  all_results[[pg]] <- page_data$results
  if (nrow(page_data$results) < 100) break
}

spending_data <- page_data  # keep metadata from last response
spending_data$results <- bind_rows(all_results)

# Extract the results into a tidy dataframe
awards <- spending_data$results |>
  as_tibble() |>
  mutate(
    award_amount = as.numeric(`Award Amount`),
    # Build a clickable link to the award page on USAspending.gov
    Award = ifelse(
      !is.na(generated_internal_id) & generated_internal_id != "",
      paste0("[", `Award ID`, "](https://www.usaspending.gov/award/",
             generated_internal_id, ")"),
      `Award ID`
    )
  ) |>
  arrange(desc(award_amount))

cat(
  "Total matching awards:", spending_data$page_metadata$total, "\n",
  "Awards retrieved:", nrow(awards), "\n",
  "Total value of retrieved awards: $",
  format(sum(awards$award_amount, na.rm = TRUE), big.mark = ","), "\n"
)

# Truncate descriptions to a readable length for the table
awards <- awards |>
  mutate(
    Description = ifelse(
      nchar(Description) > 120,
      paste0(substr(Description, 1, 120), "..."),
      Description
    )
  )

# Display the top 5 awards by amount
awards |>
  select(
    Award, `Recipient Name`, `Award Amount`,
    `Awarding Sub Agency`, Description, `Start Date`
  ) |>
  head(5) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = paste0("Top 5 Opioid-Related Federal Grant Awards by Amount (",
                     year_start, "-", year_end, ")")
  )
```

### What Are These Awards Funding?

The word cloud below visualizes the most frequent terms appearing in award descriptions across all `r nrow(awards)` retrieved opioid-related grants. Larger words appear more often, revealing the dominant themes in federal spending.

```{r}
#| label: usaspending-wordcloud
#| eval: true
#| code-fold: true

library(tidytext)
library(wordcloud2)

# Use the full (untruncated) descriptions from the API response
full_descriptions <- spending_data$results$Description

# Tokenize into individual words, remove stop words and generic terms
word_counts <- tibble(text = full_descriptions) |>
  filter(!is.na(text), text != "") |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  # Remove generic/uninformative terms and numbers
  filter(
    !word %in% c("award", "grant", "funding", "program", "provide",
                  "project", "based", "including", "support", "services",
                  "funds", "federal", "national", "department", "health",
                  "1", "2", "3", "4", "5", "01", "02", "12",
                  as.character(year_start:year_end), "cfda", "fy"),
    nchar(word) > 2,
    !grepl("^[0-9]+$", word)
  ) |>
  count(word, sort = TRUE)

# Display word cloud using top 100 terms
wordcloud2(
  word_counts |> head(100),
  size = 0.6,
  color = rep_len(
    c("#2494f7", "#00a4bb", "#01272f", "#204d70"),
    100
  ),
  backgroundColor = "white",
  fontFamily = "Inter",
  shape = "circle"
)
```

### Federal Funding Over Time

The chart below plots individual grant award amounts by their start date, colored by awarding sub-agency. Hover over each point to see the recipient and award details.

```{r}
#| label: usaspending-funding-timeline
#| eval: true
#| code-fold: true

# Prepare data for the timeline chart
funding_timeline <- awards |>
  mutate(
    start_date = as.Date(`Start Date`),
    award_amount = as.numeric(`Award Amount`),
    agency = `Awarding Sub Agency`,
    recipient = `Recipient Name`,
    hover_text = paste0(
      "<b>", agency, "</b><br>",
      "Recipient: ", recipient, "<br>",
      "Award: $", formatC(award_amount, format = "f", digits = 0, big.mark = ",")
    )
  ) |>
  filter(!is.na(start_date), !is.na(award_amount))

# Brand-aligned color palette for agencies
agency_colors <- c(
  "#2494f7", "#00a4bb", "#01272f", "#204d70",
  "#e07b39", "#6b4c9a", "#d94f4f", "#2ca02c",
  "#8c564b", "#17becf", "#bcbd22", "#7f7f7f"
)

funding_plot <- plot_ly(
  data = funding_timeline,
  x = ~start_date,
  y = ~award_amount,
  color = ~agency,
  colors = agency_colors,
  type = "scatter",
  mode = "markers",
  marker = list(size = 8, opacity = 0.7),
  text = ~hover_text,
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = paste0("Opioid-Related Federal Grant Awards Over Time (",
                    year_start, "-", year_end, ")"),
      font = list(size = 16)
    ),
    xaxis = list(
      title = "Award Start Date",
      type = "date"
    ),
    yaxis = list(
      title = "Award Amount ($)",
      separatethousands = TRUE,
      type = "log"
    ),
    legend = list(
      title = list(text = "Awarding Sub Agency"),
      orientation = "h",
      y = -0.25,
      x = 0,
      font = list(size = 9)
    ),
    hovermode = "closest",
    margin = list(b = 120)
  ) |>
  config(displayModeBar = TRUE)

funding_plot
```

The table below lists all retrieved awards with full descriptions. Use the search box to filter by keyword and click column headers to sort.

```{r}
#| label: usaspending-award-table
#| eval: true
#| code-fold: true

library(DT)

awards |>
  mutate(
    `Start Date` = as.Date(`Start Date`),
    `Award Amount` = as.numeric(`Award Amount`)
  ) |>
  select(
    `Awarding Sub Agency`, `Recipient Name`, `Award Amount`,
    `Start Date`, Description
  ) |>
  arrange(desc(`Award Amount`)) |>
  datatable(
    caption = paste0("Opioid-Related Federal Grant Awards (",
                     year_start, "-", year_end, ")"),
    filter = "top",
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(2, "desc")),
      columnDefs = list(
        list(width = "300px", targets = 4)
      )
    ),
    rownames = FALSE
  ) |>
  formatCurrency("Award Amount", digits = 0) |>
  formatDate("Start Date", method = "toLocaleDateString")
```

::: {.callout-note title="Data Source"}
USAspending is the official U.S. government open data source for federal spending. HHS TAGGS provides more granular transaction-level tracking of HHS grant obligations specifically.
:::

## Research Literature and Databases

Beyond surveillance data, a large and growing body of peer-reviewed research examines the epidemiology, pharmacology, policy responses, and social determinants of the overdose crisis. Several public databases allow programmatic access to this literature and to information about funded research projects.

### PubMed (NCBI E-utilities API)

[PubMed](https://pubmed.ncbi.nlm.nih.gov/) indexes over 37 million biomedical citations. The [NCBI E-utilities API](https://www.ncbi.nlm.nih.gov/books/NBK25501/) enables programmatic search and retrieval without authentication (subject to rate limits of 3 requests per second without an API key, or 10 per second with one).

```{r}
#| label: pubmed-search
#| eval: false
#| code-fold: false

library(httr2)
library(jsonlite)
library(dplyr)
library(tidyr)

# NCBI E-utilities: search PubMed for recent opioid overdose research
# Documentation: https://www.ncbi.nlm.nih.gov/books/NBK25501/

# Step 1: Search for article IDs matching our query
esearch_url <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi"

search_response <- request(esearch_url) |>
  req_url_query(
    db       = "pubmed",
    term     = "opioid overdose fentanyl United States[Title/Abstract]",
    retmax   = 20,
    sort     = "date",
    retmode  = "json",
    mindate  = format(date_start, "%Y/%m/%d"),
    maxdate  = format(date_end, "%Y/%m/%d"),
    datetype = "pdat"
  ) |>
  req_perform()

search_results <- search_response |>
  resp_body_json()

# Extract PubMed IDs
pmids <- search_results$esearchresult$idlist
cat("Found", search_results$esearchresult$count, "articles.\n")
cat("Retrieved", length(pmids), "PubMed IDs.\n")

# Step 2: Fetch article summaries for those IDs
esummary_url <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esummary.fcgi"

if (length(pmids) > 0) {
  summary_response <- request(esummary_url) |>
    req_url_query(
      db      = "pubmed",
      id      = paste(pmids, collapse = ","),
      retmode = "json"
    ) |>
    req_perform()

  summaries <- summary_response |>
    resp_body_json()

  # Parse into a tidy dataframe
  articles <- lapply(pmids, function(id) {
    article <- summaries$result[[id]]
    tibble(
      pmid      = id,
      title     = article$title %||% NA_character_,
      journal   = article$fulljournalname %||% NA_character_,
      pub_date  = article$pubdate %||% NA_character_,
      doi       = {
        doi_ids <- Filter(function(x) x$idtype == "doi", article$articleids)
        if (length(doi_ids) > 0) doi_ids[[1]]$value else NA_character_
      }
    )
  }) |>
    bind_rows()

  print(articles)
}
```

::: {.callout-tip title="Reproducibility"}
For higher rate limits, register for a free NCBI API key at [ncbi.nlm.nih.gov/account](https://www.ncbi.nlm.nih.gov/account/) and append `&api_key=YOUR_KEY` to requests. The E-utilities support complex Boolean queries using MeSH terms (e.g., `"Opiate Overdose"[MeSH] AND "Fentanyl"[MeSH]`) for more precise results.
:::

### NIH RePORTER (Funded Research Projects)

[NIH RePORTER](https://reporter.nih.gov/) tracks all NIH-funded research projects, including active grants, publications, patents, and clinical trials. Its [API](https://api.reporter.nih.gov/) supports POST-based queries to identify what research is currently being funded on any topic.

```{r}
#| label: nih-reporter-search
#| eval: false
#| code-fold: false

# NIH RePORTER API: find funded opioid overdose research projects
# Documentation: https://api.reporter.nih.gov/

reporter_url <- "https://api.reporter.nih.gov/v2/projects/search"

reporter_body <- list(
  criteria = list(
    advanced_text_search = list(
      operator = "and",
      search_field = "projecttitle,terms",
      search_text = "opioid overdose fentanyl"
    ),
    fiscal_years = as.list((year_end - 2L):year_end),
    exclude_subprojects = TRUE
  ),
  offset = 0,
  limit = 25,
  sort_field = "FiscalYear",
  sort_order = "desc"
)

reporter_response <- request(reporter_url) |>
  req_body_json(reporter_body) |>
  req_perform()

reporter_data <- reporter_response |>
  resp_body_json()

cat("Total matching projects:", reporter_data$meta$total, "\n")

# Parse results into a dataframe
if (length(reporter_data$results) > 0) {
  projects <- lapply(reporter_data$results, function(p) {
    tibble(
      project_num   = p$project_num %||% NA_character_,
      fiscal_year   = p$fiscal_year %||% NA_integer_,
      title         = p$project_title %||% NA_character_,
      pi_name       = {
        if (!is.null(p$principal_investigators) && length(p$principal_investigators) > 0)
          p$principal_investigators[[1]]$full_name
        else NA_character_
      },
      organization  = p$organization$org_name %||% NA_character_,
      award_amount  = p$award_amount %||% NA_real_,
      abstract_text = substr(p$abstract_text %||% "", 1, 200)
    )
  }) |>
    bind_rows()

  print(projects |> select(project_num, fiscal_year, title, award_amount))
}
```

::: {.callout-note title="Data Source"}
NIH RePORTER is the authoritative source for federally funded biomedical research. It covers grants from NIH, CDC, FDA, AHRQ, SAMHSA, and other HHS agencies. No authentication is required. The API returns project details, funding amounts, and links to resulting publications.
:::

### Europe PMC (Open Access Literature)

[Europe PMC](https://europepmc.org/) provides open access to a worldwide collection of life science publications, including all PubMed content plus preprints, patents, and clinical guidelines. Its [REST API](https://europepmc.org/RestfulWebService) is particularly useful for retrieving full-text open access articles.

```{r}
#| label: europe-pmc-search
#| eval: false
#| code-fold: false

# Europe PMC REST API: search for open access opioid research
# Documentation: https://europepmc.org/RestfulWebService

epmc_url <- "https://www.ebi.ac.uk/europepmc/webservices/rest/search"

epmc_response <- request(epmc_url) |>
  req_url_query(
    query      = paste0("(opioid overdose fentanyl) AND (PUB_YEAR:[",
                        year_start, " TO ", year_end, "])"),
    format     = "json",
    pageSize   = 15,
    sort       = "date",
    resultType = "core"
  ) |>
  req_perform()

epmc_data <- epmc_response |>
  resp_body_json()

cat("Total results:", epmc_data$hitCount, "\n")

if (length(epmc_data$resultList$result) > 0) {
  epmc_articles <- lapply(epmc_data$resultList$result, function(a) {
    tibble(
      pmid            = a$pmid %||% NA_character_,
      title           = a$title %||% NA_character_,
      journal         = a$journalTitle %||% NA_character_,
      year            = a$pubYear %||% NA_character_,
      is_open_access  = a$isOpenAccess %||% "N",
      cited_by_count  = a$citedByCount %||% 0L,
      doi             = a$doi %||% NA_character_
    )
  }) |>
    bind_rows()

  # Show most-cited open access articles
  epmc_articles |>
    filter(is_open_access == "Y") |>
    arrange(desc(cited_by_count)) |>
    print()
}
```

### ClinicalTrials.gov (Clinical Research)

[ClinicalTrials.gov](https://clinicaltrials.gov/) registers clinical studies on a wide range of conditions, including opioid use disorder treatment trials, naloxone distribution studies, and harm reduction interventions. The [v2 API](https://clinicaltrials.gov/data-api/about-api) supports structured queries.

```{r}
#| label: clinicaltrials-search
#| eval: false
#| code-fold: false

# ClinicalTrials.gov v2 API: find opioid-related clinical trials
# Documentation: https://clinicaltrials.gov/data-api/about-api

ct_url <- "https://clinicaltrials.gov/api/v2/studies"

ct_response <- request(ct_url) |>
  req_url_query(
    `query.cond`  = "opioid overdose",
    `query.intr`  = "fentanyl OR naloxone OR buprenorphine",
    `filter.overallStatus` = "RECRUITING,ACTIVE_NOT_RECRUITING,COMPLETED",
    pageSize      = 15,
    sort          = "LastUpdatePostDate:desc",
    fields        = "NCTId,BriefTitle,OverallStatus,StartDate,CompletionDate,LeadSponsorName,EnrollmentCount"
  ) |>
  req_perform()

ct_data <- ct_response |>
  resp_body_json()

cat("Total studies found:", ct_data$totalCount, "\n")

if (length(ct_data$studies) > 0) {
  trials <- lapply(ct_data$studies, function(s) {
    proto <- s$protocolSection
    id_mod <- proto$identificationModule
    status_mod <- proto$statusModule
    sponsor_mod <- proto$sponsorCollaboratorsModule
    design_mod <- proto$designModule

    tibble(
      nct_id     = id_mod$nctId %||% NA_character_,
      title      = id_mod$briefTitle %||% NA_character_,
      status     = status_mod$overallStatus %||% NA_character_,
      start_date = status_mod$startDateStruct$date %||% NA_character_,
      sponsor    = sponsor_mod$leadSponsor$name %||% NA_character_,
      enrollment = design_mod$enrollmentInfo$count %||% NA_integer_
    )
  }) |>
    bind_rows()

  print(trials)
}
```

::: {.callout-tip title="Reproducibility"}
ClinicalTrials.gov does not require authentication. The v2 API returns JSON by default and supports filtering by condition, intervention, status, location, and date range. This is useful for mapping the research landscape around specific interventions (e.g., naloxone distribution, buprenorphine access).
:::

### Global Burden of Disease (GBD) Results Tool

The [IHME GBD Results Tool](https://vizhub.healthdata.org/gbd-results/) provides modeled estimates for drug use disorders as a specific cause of death and disability across 204 countries. This is the authoritative source for comparing the U.S. overdose burden against other nations using standardized DALY, YLL, and YLD metrics. Data can be exported as CSV after creating a free account at [healthdata.org](https://www.healthdata.org/research-analysis/gbd-data).

::: {.callout-note title="Data Source"}
The GBD study models both fatal and non-fatal outcomes of drug use disorders, enabling analysis that goes beyond death counts. This captures the full burden, including years lived with disability from substance use disorders, which is not available in vital statistics data alone.
:::

## Questions for Further Analysis

This chapter provides the starting framework. Future iterations could explore:

- **Trend analysis**: How have overdose death rates changed year-over-year by state and by substance type?
- **Demographic patterns**: Which age groups, races, and urban/rural classifications bear the highest burden?
- **Supply vs. demand**: How do geographic overdose patterns correlate with drug trafficking routes?
- **Spending alignment**: Do states with higher overdose rates receive proportionally more federal prevention funding?
- **County-level mapping**: Where within states are deaths concentrated?

- **Research landscape**: What does the peer-reviewed literature reveal about effective interventions? Which NIH-funded projects are producing results?
- **International comparison**: How does the U.S. drug use disorder DALY rate compare against peer nations using GBD estimates?

Each of these questions is answerable with publicly available data, research databases, and the analytical patterns demonstrated above.

# References