---
title: "The U.S. Drug Overdose Crisis"
author: "André van Zyl, MPH"
---

# The U.S. Drug Overdose Crisis

The United States stands alone among high-income nations in the scale of its overdose crisis. Drug overdoses claimed more than 100,000 lives for a third consecutive year in 2023, representing over a 50 percent increase since 2019 [@commonwealth_overdose_intl]. With an overdose rate of 324 deaths per million people (nearly 108,000 deaths in 2022), the U.S. has by far the highest overdose death rate in the world. While Scotland and Canada, the next highest-ranked countries, saw declines from 2021 to 2022, rates in the U.S. have remained persistently elevated [@commonwealth_overdose_intl].

According to the [CDC National Center for Health Statistics](https://www.cdc.gov/nchs/fastats/leading-causes-of-death.htm), the three leading causes of death in the U.S. are heart disease (680,981 deaths), cancer (613,352), and unintentional injuries (222,698), with drug poisonings dominating the latter category [@cdc_faststats]. Drug overdose has become a [leading cause of injury mortality](https://www.cdc.gov/nchs/hus/topics/drug-overdose-deaths.htm), with opioids driving an increasing share of these deaths since 2009 and accounting for over 70% of all overdose fatalities [@cdc_hus_overdose]. The growing involvement of stimulants such as cocaine and methamphetamine alongside opioids has further complicated the crisis, transforming what began as a prescription opioid problem into a polysubstance mass-casualty event.

No peer nation experiences anything comparable. The [Commonwealth Fund 2025 Scorecard on State Health System Performance](https://www.commonwealthfund.org/publications/scorecard/2025/jun/2025-scorecard-state-health-system-performance) documents that premature, avoidable deaths are increasing in the United States while declining in comparable high-income countries, with drug overdose a major contributor to this divergence [@commonwealth2025]. The rate of avoidable mortality in West Virginia is more than twice the rate in Massachusetts, and wide racial disparities in premature deaths are the norm in most states [@commonwealth2025]. While overdose deaths fell slightly between 2022 and 2023 in 37 states, the total remains above 100,000 annually, and rates spiked by 25% or more in Nevada, Oregon, Washington, and Alaska [@commonwealth2025].

This chapter explores what the data shows, where the crisis hits hardest, and how to access and analyze the underlying datasets and research literature programmatically.

## Why This Chapter

My path to this topic stems from my educational background and professional experience. I studied psychology and counseling in South Africa, where I worked as a supportive counselor at a residential recovery center for people with substance use disorders. That experience gave me a ground-level understanding of addiction, recovery, and the human cost of substance-related harm.

When I moved to the United States, I helped with the development of COVID-19 contact tracing systems before joining a local health department as an epidemiologist, where my responsibilities included drug overdose mortality monitoring and surveillance. I worked with county-level data coming from different jurisdictions, most of which was not programmatically accessible. After working at the health department, I continued at the CDC on COVID-19 response work and then antimicrobial resistance (AMR) response, but the opioid epidemic never left my peripheral vision.

Now, with more capacity to investigate, I am returning to the overdose landscape with fresh analytical tools and a broader perspective. As the published literature documents, the scale of drug overdose mortality in the United States is without parallel among high-income nations [@commonwealth2025]. This chapter is my attempt to understand the current state of the crisis: to see what data is available, to assess the impact and scale, to trace where funding is flowing, to examine what interventions are being tried, and to survey the research being conducted in the field.

## The Scale of the Problem

Drug overdose deaths in the United States have followed a steep, accelerating trajectory over the past 25 years. Approximately 105,000 people died from drug overdose in 2023, with nearly 80,000 (76%) of those deaths involving opioids. The number of opioid overdose deaths in 2023 was nearly ten times the number in 1999, though 2023 marked the first annual decline since 2018 [@cdc_opioid_epidemic]. From 1999 to 2023, approximately 806,000 people died from an opioid overdose in the United States [@cdc_opioid_epidemic].

The [CDC describes three distinct waves](https://www.cdc.gov/overdose-prevention/about/understanding-the-opioid-overdose-epidemic.html) of opioid overdose deaths [@cdc_opioid_epidemic]:

1. **Wave 1 (1990s)**: Increased prescribing of opioids drove a rise in overdose deaths involving prescription opioids (natural and semi-synthetic opioids and methadone) starting around 1999, though these have declined in recent years
2. **Wave 2 (2010)**: Rapid increases in overdose deaths involving heroin began in 2010, though heroin overdose deaths have also been declining in recent years
3. **Wave 3 (2013 onward)**: Substantial increases in overdose deaths involving synthetic opioids, particularly illegally manufactured fentanyls (IMFs), which have largely saturated the illegal drug supply

The third wave remains the dominant driver. IMFs are found in powder form or pressed into counterfeit pills and frequently contain other drugs, including non-opioid sedatives such as xylazine [@cdc_opioid_epidemic]. From 2022 to 2023, the rate of overdose deaths involving synthetic opioids decreased approximately 2%, heroin deaths decreased roughly 33%, and prescription opioid deaths decreased nearly 12% [@cdc_opioid_epidemic]. Despite these declines, in 2023, nearly 47% of drug overdose deaths involved both opioids and stimulants, underscoring the polysubstance nature of the current crisis [@cdc_opioid_epidemic]. Data from the [SUDORS Dashboard](https://www.cdc.gov/overdose-prevention/data-research/facts-stats/sudors-dashboard-fatal-overdose-data.html) further confirms this pattern: in 2024, 73.4% of overdose deaths involved at least one opioid and 65.1% involved at least one stimulant [@cdc_sudors]. Critically, 67.5% of overdose deaths had at least one potential opportunity for intervention, and in 2022, 77% of adults with substance use disorder were not receiving treatment [@commonwealth2025].

## Data Sources

The primary datasets for analyzing drug overdose mortality in the United States are:

### CDC WONDER

[CDC WONDER](https://wonder.cdc.gov/) (Wide-ranging ONline Data for Epidemiologic Research) is the premier federal system for querying mortality data from the National Vital Statistics System. It provides:

- Mortality counts, crude rates, and age-adjusted rates
- Stratification by geography (national, state, county), demographics, urbanization, and ICD-10 codes
- Multiple Cause of Death files that capture all substances listed on the death certificate

::: {.callout-warning title="Data Suppression"}
CDC WONDER suppresses counts below 10 deaths in any given cell to prevent identification. Analyses of small populations or rare drug categories at the county level will encounter this limitation.
:::

### CDC Open Data (WONDER API and Socrata)

CDC publishes several overdose-specific datasets on its open data portal at [data.cdc.gov](https://data.cdc.gov/), including provisional drug overdose death counts updated monthly. These datasets are Socrata-hosted and accessible via API.

### State and Local Portals

Many states and local jurisdictions maintain their own open data portals with overdose-related datasets, often including more granular geographic or toxicological detail than federal sources. These vary in format, update frequency, and accessibility, but are worth exploring for jurisdiction-specific analyses.

## Accessing the Data with R

### VSRR Provisional Drug Overdose Death Counts

The primary dataset used in this chapter is the [VSRR Provisional Drug Overdose Death Counts](https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a) (dataset ID: `xkb8-kh2a`), published by the [National Center for Health Statistics (NCHS)](https://www.cdc.gov/nchs/nvss/vsrr/drug-overdose-data.htm) as part of the Vital Statistics Rapid Release (VSRR) program. This dataset presents provisional counts of drug overdose deaths based on the current flow of mortality data in the [National Vital Statistics System (NVSS)](https://www.cdc.gov/nchs/nvss/index.htm). It is updated monthly, with data available within four months after the date of death.

Provisional counts may not include all deaths that occurred during a given period because drug overdose deaths often require lengthy investigation, including toxicology testing. Death certificates may initially be filed with manner of death "pending investigation" or with a preliminary cause of death. To address this undercount, NCHS developed statistical methods to generate **predicted provisional counts** that adjust for reporting delays based on historical completeness patterns. Drug overdose deaths are identified using ICD-10 underlying cause-of-death codes: X40--X44 (unintentional), X60--X64 (suicide), X85 (homicide), and Y10--Y14 (undetermined intent).

The dataset reports **12-month ending counts**, meaning each row represents the total number of deaths occurring in the 12-month period ending in the month indicated. For example, a row for June 2024 captures deaths from July 2023 through June 2024. This rolling window smooths out seasonal variation and provides a stable trend measure. The `indicator` field specifies which outcome is being counted: total drug overdose deaths, or deaths involving a specific drug or drug class, identified by ICD-10 multiple cause-of-death codes:

- **Heroin** (T40.1)
- **Natural opioid analgesics** including morphine, codeine, oxycodone, hydrocodone (T40.2)
- **Methadone** (T40.3)
- **Synthetic opioids other than methadone**, including fentanyl and tramadol (T40.4)
- **Cocaine** (T40.5)
- **Psychostimulants with abuse potential**, including methamphetamine (T43.6)

Because a single death may involve multiple substances, it can appear in more than one drug category. For example, a death involving both heroin and fentanyl would be counted under both heroin and synthetic opioids other than methadone.

#### Dataset Columns

The dataset contains 12 fields:

| Column | API Field | Type | Description |
|:---|:---|:---|:---|
| State | `state` | Text | Two-letter jurisdiction abbreviation (e.g., "US", "CO", "NM") |
| Year | `year` | Number | Calendar year of the 12-month ending period |
| Month | `month` | Text | Month name of the 12-month ending period (e.g., "January") |
| Period | `period` | Text | Label describing the time span (e.g., "12 Month-ending Provisional Counts") |
| Indicator | `indicator` | Text | The specific outcome measured: "Number of Drug Overdose Deaths" or a drug-specific category such as "Heroin (T40.1)" or "Synthetic opioids, excl. methadone (T40.4)" |
| Data Value | `data_value` | Number | The reported provisional count of deaths for the indicator, jurisdiction, and period |
| Predicted Value | `predicted_value` | Number | The model-adjusted count that corrects for reporting delays and pending investigations |
| Percent Complete | `percent_complete` | Text | Percentage of expected death records received and processed for the jurisdiction and period |
| Percent Pending Investigation | `percent_pending_investigation` | Number | Percentage of death records where manner of death is listed as "pending investigation" |
| State Name | `state_name` | Text | Full jurisdiction name (e.g., "United States", "Colorado") |
| Footnote | `footnote` | Text | Contextual notes about data quality or suppression for the observation |
| Footnote Symbol | `footnote_symbol` | Text | Symbol corresponding to the footnote (if any) |

::: {.callout-warning title="Interpreting Provisional Data"}
Provisional counts are often lower than final counts due to reporting delays. The `predicted_value` column provides a corrected estimate but may not fully account for unusually long delays in some jurisdictions. Comparisons of drug-specific death counts across states should be made cautiously because the percentage of death certificates with specific drugs recorded varies by jurisdiction. For definitive counts, use the [final mortality data](https://www.cdc.gov/nchs/nvss/mortality_public_use_data.htm) published annually through CDC WONDER.
:::

The following code demonstrates how to pull this dataset from CDC's Socrata-hosted open data portal.

```{r}
#| label: setup
#| include: false

# Load all libraries upfront so they remain available even when
# downstream data-fetching chunks are loaded from knitr cache
# (cached chunks do not re-execute library() calls).
library(httr2)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(tidytext)
library(ggwordcloud)
library(tidyr)
library(knitr)
```

```{r}
#| label: cdc-overdose-data
#| eval: true
#| cache: true
#| cache.extra: !expr Sys.Date()
#| code-fold: true

# CDC provisional drug overdose death counts (Socrata endpoint)
# Dataset: VSRR Provisional Drug Overdose Death Counts
# Source: https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a

cdc_url <- "https://data.cdc.gov/resource/xkb8-kh2a.json"

# Query: US + all states, total drug overdose deaths only
# The Socrata API caps at 50,000 rows per request
response <- request(cdc_url) |>
  req_url_query(
    `$where`  = "indicator = 'Number of Drug Overdose Deaths'",
    `$limit`  = 50000,
    `$order`  = "year ASC, month ASC"
  ) |>
  req_perform()

overdose_raw <- response |>
  resp_body_json(simplifyVector = TRUE)
```

The query returned **`r nrow(overdose_raw)`** rows covering **`r length(unique(overdose_raw$state_name))`** jurisdictions.

::: {.callout-tip title="Reproducibility"}
The Socrata API does not require authentication for public datasets. The dataset identifier `xkb8-kh2a` is stable. You can also download the full dataset as CSV from the [dataset page](https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a). For full methodology, including the statistical models used to generate predicted counts and the data quality thresholds applied to each jurisdiction, see the [Technical Notes](https://www.cdc.gov/nchs/nvss/vsrr/drug-overdose-data.htm).
:::

#### Interactive Time Series: Drug Overdose Deaths by Jurisdiction

The plot below shows 12-month ending provisional drug overdose death counts for the United States and every reporting state. Use the legend to toggle individual jurisdictions on or off, hover over data points for exact counts, and use the range slider to zoom into specific time periods.

```{r}
#| label: overdose-time-series
#| eval: true
#| code-fold: true

# Month lookup for proper date ordering
month_lookup <- c(
  "January" = 1, "February" = 2, "March" = 3, "April" = 4,
  "May" = 5, "June" = 6, "July" = 7, "August" = 8,
  "September" = 9, "October" = 10, "November" = 11, "December" = 12
)

overdose_ts <- overdose_raw |>
  mutate(
    data_value     = as.numeric(data_value),
    predicted_value = as.numeric(predicted_value),
    year_num       = as.integer(year),
    month_num      = month_lookup[month],
    date           = as.Date(paste(year_num, month_num, "01", sep = "-"))
  ) |>
  filter(!is.na(data_value)) |>
  arrange(state_name, date)

# Separate United States (always visible) from states (hidden by default)
us_data    <- overdose_ts |> filter(state_name == "United States")
state_data <- overdose_ts |> filter(state_name != "United States")

# Get sorted list of state names
state_names <- sort(unique(state_data$state_name))

# Build the plotly figure
p <- plot_ly() |>
  # United States trend line (bold, always visible)
  add_trace(
    data = us_data,
    x = ~date, y = ~data_value,
    type = "scatter", mode = "lines",
    name = "United States",
    line = list(color = "#2494f7", width = 3),
    hovertemplate = paste(
      "<b>United States</b><br>",
      "Period ending: %{x|%B %Y}<br>",
      "Reported deaths: %{y:,.0f}<extra></extra>"
    )
  )

# Add each state as a separate trace (hidden by default via legendgroup)
for (st in state_names) {
  st_df <- state_data |> filter(state_name == st)
  p <- p |>
    add_trace(
      data = st_df,
      x = ~date, y = ~data_value,
      type = "scatter", mode = "lines",
      name = st,
      visible = "legendonly",
      hovertemplate = paste0(
        "<b>", st, "</b><br>",
        "Period ending: %{x|%B %Y}<br>",
        "Reported deaths: %{y:,.0f}<extra></extra>"
      )
    )
}

p <- p |>
  layout(
    title = list(
      text = "12-Month Ending Provisional Drug Overdose Deaths by Jurisdiction",
      font = list(size = 16)
    ),
    xaxis = list(
      title = "12-Month Period Ending",
      rangeslider = list(visible = TRUE),
      type = "date"
    ),
    yaxis = list(
      title = "Deaths (12-month ending count)",
      separatethousands = TRUE
    ),
    legend = list(
      orientation = "v",
      x = 1.02, y = 1,
      font = list(size = 10)
    ),
    hovermode = "x unified",
    margin = list(r = 150)
  ) |>
  config(displayModeBar = TRUE)

p
```

#### Overdose Data Detail

The table below lists the full CDC provisional overdose dataset. Use the filters to explore specific states, years, or months.

```{r}
#| label: overdose-data-table
#| eval: true
#| code-fold: true

library(DT)

overdose_raw |>
  mutate(
    data_value = as.numeric(data_value),
    predicted_value = as.numeric(predicted_value),
    percent_complete = as.numeric(percent_complete)
  ) |>
  select(
    `State Name` = state_name, Year = year, Month = month,
    `Reported Deaths` = data_value, `Predicted Deaths` = predicted_value,
    `% Complete` = percent_complete
  ) |>
  datatable(
    caption = "CDC VSRR Provisional Drug Overdose Death Counts",
    filter = "top",
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(1, "asc"), list(2, "desc"))
    ),
    rownames = FALSE
  ) |>
  formatRound(c("Reported Deaths", "Predicted Deaths"), digits = 0) |>
  formatRound("% Complete", digits = 1)
```

## What the Interventions Look Like

Public health responses to the overdose crisis operate at multiple levels:

**Harm reduction**: Expanding naloxone distribution (the opioid reversal drug), establishing syringe services programs, and drug checking services [@cdc_overdose_prevention]. Colorado has enacted harm reduction legislation expanding naloxone access via standing orders through pharmacies; the [Colorado Consortium for Prescription Drug Abuse Prevention](https://corxconsortium.org/data/) tracks the state's comprehensive response [@corxconsortium]. New Mexico operates state-funded [harm reduction programs](https://www.nmhealth.org/about/phd/idb/hrp/) focused on overdose death reduction, overdose prevention training, naloxone distribution, and participant education [@nmdoh_harm_reduction].

**Treatment access**: Increasing access to medications for opioid use disorder (MOUD), including buprenorphine, methadone, and naltrexone [@cdc_overdose_prevention]. Federal funding through State Opioid Response (SOR) grants ($1.48 billion nationally) and Tribal Opioid Response grants (~$63 million) have supported state and local treatment programs. However, the reduced footprint of the Substance Abuse and Mental Health Services Administration (SAMHSA) and the loss of roughly $11 billion in grants that had been supporting state and local substance use treatment programs threaten to reverse recent progress [@commonwealth2025].

**Surveillance**: Enhanced death investigation through SUDORS (State Unintentional Drug Overdose Reporting System) provides toxicological detail beyond what death certificates capture, revealing which specific substances are involved [@cdc_sudors].

## Federal Funding

Opioid response funding flows through several federal mechanisms.

### USAspending

The [USAspending API](https://api.usaspending.gov/) enables programmatic tracking of federal grant allocations related to the opioid crisis:

```{r}
#| label: usaspending-opioid
#| eval: true
#| cache: true
#| cache.extra: !expr Sys.Date()
#| code-fold: true

# Query USAspending for opioid-related grants
# Documentation: https://api.usaspending.gov/

usa_spending_url <- "https://api.usaspending.gov/api/v2/search/spending_by_award/"

# Dynamic date window: past 5 years from today
date_end   <- Sys.Date()
date_start <- as.Date(paste0(as.integer(format(date_end, "%Y")) - 5, "-01-01"))
year_end   <- as.integer(format(date_end, "%Y"))
year_start <- year_end - 5

# POST request body: opioid-related keyword search for grants
request_body <- list(
  filters = list(
    keywords = list("opioid", "overdose", "substance abuse"),
    time_period = list(
      list(start_date = format(date_start, "%Y-%m-%d"),
           end_date   = format(date_end, "%Y-%m-%d"))
    ),
    award_type_codes = c("02", "03", "04", "05")
  ),
  fields = c(
    "Award ID", "Recipient Name", "Award Amount",
    "Awarding Agency", "Awarding Sub Agency",
    "Start Date", "End Date", "Description",
    "generated_internal_id"
  ),
  limit = 100,
  page = 1
)

# Paginate to collect all available awards
# USAspending API caps at 100 results per page
all_results <- list()
for (pg in 1:100) {
  request_body$page <- pg
  spending_response <- request(usa_spending_url) |>
    req_body_json(request_body) |>
    req_perform()

  page_data <- spending_response |>
    resp_body_json(simplifyVector = TRUE)

  if (is.null(page_data$results) || nrow(page_data$results) == 0) break
  all_results[[pg]] <- page_data$results
  if (nrow(page_data$results) < 100) break
}

spending_data <- page_data  # keep metadata from last response
spending_data$results <- bind_rows(all_results)

# Extract the results into a tidy dataframe
awards <- spending_data$results |>
  as_tibble() |>
  mutate(
    award_amount = as.numeric(`Award Amount`),
    # Build a clickable link to the award page on USAspending.gov
    Award = ifelse(
      !is.na(generated_internal_id) & generated_internal_id != "",
      paste0("[", `Award ID`, "](https://www.usaspending.gov/award/",
             generated_internal_id, ")"),
      `Award ID`
    )
  ) |>
  arrange(desc(award_amount))

# Truncate descriptions to a readable length for the table
awards <- awards |>
  mutate(
    Description = ifelse(
      nchar(Description) > 120,
      paste0(substr(Description, 1, 120), "..."),
      Description
    )
  )

# Display the top 5 awards by amount
awards |>
  select(
    Award, `Recipient Name`, `Award Amount`,
    `Awarding Sub Agency`, Description, `Start Date`
  ) |>
  head(5) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = paste0("Top 5 Opioid-Related Federal Grant Awards by Amount (",
                     year_start, "-", year_end, ")")
  )
```

The query returned **`r format(nrow(awards), big.mark = ",")`** awards, with a combined value of **$`r format(sum(awards$award_amount, na.rm = TRUE), big.mark = ",")`**.

#### What Are These Awards Funding?

The word cloud below visualizes the most frequent terms appearing in award descriptions across all `r format(nrow(awards), big.mark = ",")` retrieved opioid-related grants. Larger words appear more often, revealing the dominant themes in federal spending.

```{r}
#| label: usaspending-wordcloud
#| eval: true
#| code-fold: true
#| column: page
#| fig-width: 14
#| fig-height: 4
#| out-width: "100%"

# Use the full (untruncated) descriptions from the API response
full_descriptions <- spending_data$results$Description

# Tokenize into individual words, remove stop words and generic terms
word_counts <- tibble(text = full_descriptions) |>
  filter(!is.na(text), text != "") |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(
    !word %in% c("award", "grant", "funding", "program", "provide",
                  "project", "based", "including", "support", "services",
                  "funds", "federal", "national", "department", "health",
                  "1", "2", "3", "4", "5", "01", "02", "12",
                  as.character(year_start:year_end), "cfda", "fy"),
    nchar(word) > 2,
    !grepl("^[0-9]+$", word)
  ) |>
  count(word, sort = TRUE)

ggplot(word_counts |> head(80), aes(label = word, size = n, color = n)) +
  geom_text_wordcloud_area(family = "sans", shape = "square") +
  scale_size_area(max_size = 28) +
  scale_color_gradient(low = "#204d70", high = "#2494f7") +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(0, 0, 0, 0)
  )
```

#### Federal Funding Over Time

The chart below plots individual grant award amounts by their start date, colored by awarding sub-agency. Hover over each point to see the recipient and award details.

```{r}
#| label: usaspending-funding-timeline
#| eval: true
#| code-fold: true

# Prepare data for the timeline chart
funding_timeline <- awards |>
  mutate(
    start_date = as.Date(`Start Date`),
    award_amount = as.numeric(`Award Amount`),
    agency = `Awarding Sub Agency`,
    recipient = `Recipient Name`,
    hover_text = paste0(
      "<b>", agency, "</b><br>",
      "Recipient: ", recipient, "<br>",
      "Award: $", formatC(award_amount, format = "f", digits = 0, big.mark = ",")
    )
  ) |>
  filter(!is.na(start_date), !is.na(award_amount))

# Brand-aligned color palette for agencies
agency_colors <- c(
  "#2494f7", "#00a4bb", "#01272f", "#204d70",
  "#e07b39", "#6b4c9a", "#d94f4f", "#2ca02c",
  "#8c564b", "#17becf", "#bcbd22", "#7f7f7f"
)

funding_plot <- plot_ly(
  data = funding_timeline,
  x = ~start_date,
  y = ~award_amount,
  color = ~agency,
  colors = agency_colors,
  type = "scatter",
  mode = "markers",
  marker = list(size = 8, opacity = 0.7),
  text = ~hover_text,
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = paste0("Opioid-Related Federal Grant Awards Over Time (",
                    year_start, "-", year_end, ")"),
      font = list(size = 16)
    ),
    xaxis = list(
      title = "Award Start Date",
      type = "date"
    ),
    yaxis = list(
      title = "Award Amount ($)",
      separatethousands = TRUE,
      type = "log"
    ),
    legend = list(
      title = list(text = "Awarding Sub Agency"),
      orientation = "h",
      y = -0.25,
      x = 0,
      font = list(size = 9)
    ),
    hovermode = "closest",
    margin = list(b = 120)
  ) |>
  config(displayModeBar = TRUE)

funding_plot
```

#### Award Detail

The table below lists all retrieved awards with full descriptions. Use the search box to filter by keyword and click column headers to sort.

```{r}
#| label: usaspending-award-table
#| eval: true
#| code-fold: true

library(DT)

awards |>
  mutate(
    `Start Date` = as.Date(`Start Date`),
    `Award Amount` = as.numeric(`Award Amount`)
  ) |>
  select(
    `Awarding Sub Agency`, `Recipient Name`, `Award Amount`,
    `Start Date`, Description
  ) |>
  arrange(desc(`Award Amount`)) |>
  datatable(
    caption = paste0("Opioid-Related Federal Grant Awards (",
                     year_start, "-", year_end, ")"),
    filter = "top",
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(2, "desc")),
      columnDefs = list(
        list(width = "300px", targets = 4)
      )
    ),
    rownames = FALSE
  ) |>
  formatCurrency("Award Amount", digits = 0) |>
  formatDate("Start Date", method = "toLocaleDateString")
```

::: {.callout-note title="Data Source"}
USAspending is the official U.S. government open data source for federal spending. HHS TAGGS provides more granular transaction-level tracking of HHS grant obligations specifically.
:::

## Research Literature and Databases

Beyond surveillance data, a large and growing body of peer-reviewed research examines the epidemiology, pharmacology, policy responses, and social determinants of the overdose crisis. Several public databases allow programmatic access to this literature and to information about funded research projects.

### PubMed (NCBI E-utilities API)

[PubMed](https://pubmed.ncbi.nlm.nih.gov/) indexes over 37 million biomedical citations. The [NCBI E-utilities API](https://www.ncbi.nlm.nih.gov/books/NBK25501/) enables programmatic search and retrieval without authentication (subject to rate limits of 3 requests per second without an API key, or 10 per second with one).

```{r}
#| label: pubmed-search
#| eval: true
#| cache: true
#| cache.extra: !expr Sys.Date()

library(httr2)
library(jsonlite)
library(dplyr)
library(tidyr)

# NCBI E-utilities: search PubMed for recent opioid overdose research
# Documentation: https://www.ncbi.nlm.nih.gov/books/NBK25501/

# Step 1: Search for article IDs matching our query
esearch_url <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi"

search_response <- request(esearch_url) |>
  req_url_query(
    db       = "pubmed",
    term     = "opioid overdose fentanyl United States[Title/Abstract]",
    retmax   = 10000,
    sort     = "date",
    retmode  = "json",
    mindate  = format(date_start, "%Y/%m/%d"),
    maxdate  = format(date_end, "%Y/%m/%d"),
    datetype = "pdat"
  ) |>
  req_perform()

search_results <- search_response |>
  resp_body_json()

# Extract PubMed IDs
pmids <- search_results$esearchresult$idlist

# Step 2: Fetch article summaries for those IDs
esummary_url <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esummary.fcgi"

if (length(pmids) > 0) {
  summary_response <- request(esummary_url) |>
    req_url_query(
      db      = "pubmed",
      id      = paste(pmids, collapse = ","),
      retmode = "json"
    ) |>
    req_perform()

  summaries <- summary_response |>
    resp_body_json()

  # Parse into a tidy dataframe
  articles <- lapply(pmids, function(id) {
    article <- summaries$result[[id]]
    tibble(
      pmid      = id,
      title     = article$title %||% NA_character_,
      journal   = article$fulljournalname %||% NA_character_,
      pub_date  = article$pubdate %||% NA_character_,
      doi       = {
        doi_ids <- Filter(function(x) x$idtype == "doi", article$articleids)
        if (length(doi_ids) > 0) doi_ids[[1]]$value else NA_character_
      }
    )
  }) |>
    bind_rows()
}
```

The search matched **`r format(as.integer(search_results$esearchresult$count), big.mark = ",")`** articles. **`r nrow(articles)`** were retrieved for analysis.

#### Publication Trends Over Time

The chart below shows the volume of opioid overdose fentanyl research published over time. Each point represents the number of articles published in a given month.

```{r}
#| label: pubmed-timeline
#| eval: true
#| code-fold: true

library(plotly)
library(lubridate)

# Parse publication dates into proper Date objects
# PubMed dates come in formats like "2025 Jan 15", "2025 Mar", "2025"
pubmed_timeline <- articles |>
  mutate(
    pub_date_parsed = case_when(
      grepl("^\\d{4} \\w+ \\d+", pub_date) ~
        as.Date(pub_date, format = "%Y %b %d"),
      grepl("^\\d{4} \\w+", pub_date) ~
        as.Date(paste0(pub_date, " 01"), format = "%Y %b %d"),
      grepl("^\\d{4}$", pub_date) ~
        as.Date(paste0(pub_date, " Jan 01"), format = "%Y %b %d"),
      TRUE ~ NA_Date_
    ),
    pub_month = floor_date(pub_date_parsed, "month")
  ) |>
  filter(!is.na(pub_month))

# Count articles per month
monthly_counts <- pubmed_timeline |>
  count(pub_month, name = "n_articles") |>
  arrange(pub_month)

# Interactive time series
pubmed_plot <- plot_ly(
  data = monthly_counts,
  x = ~pub_month,
  y = ~n_articles,
  type = "scatter",
  mode = "lines+markers",
  line = list(color = "#2494f7", width = 2),
  marker = list(color = "#2494f7", size = 6),
  hovertemplate = paste(
    "<b>%{x|%B %Y}</b><br>",
    "Articles: %{y}<extra></extra>"
  )
) |>
  layout(
    title = list(
      text = "Opioid Overdose Fentanyl Research: Publications per Month",
      font = list(size = 16)
    ),
    xaxis = list(
      title = "Publication Month",
      type = "date",
      rangeslider = list(visible = TRUE)
    ),
    yaxis = list(
      title = "Number of Articles"
    ),
    hovermode = "x unified"
  ) |>
  config(displayModeBar = TRUE)

pubmed_plot
```

#### Research Themes in Article Titles

The word cloud below visualizes the most frequent terms appearing in the titles of all `r nrow(articles)` retrieved PubMed articles. This provides a quick thematic overview of the research landscape.

```{r}
#| label: pubmed-wordcloud
#| eval: true
#| code-fold: true
#| column: page
#| fig-width: 14
#| fig-height: 4
#| out-width: "100%"

# Tokenize article titles into individual words
pubmed_title_words <- tibble(text = articles$title) |>
  filter(!is.na(text), text != "") |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(
    !word %in% c("study", "analysis", "among", "associated",
                  "united", "states", "results", "based",
                  "findings", "using", "related", "data"),
    nchar(word) > 2,
    !grepl("^[0-9]+$", word)
  ) |>
  count(word, sort = TRUE)

ggplot(pubmed_title_words |> head(80), aes(label = word, size = n, color = n)) +
  geom_text_wordcloud_area(family = "sans", shape = "square") +
  scale_size_area(max_size = 28) +
  scale_color_gradient(low = "#204d70", high = "#2494f7") +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(0, 0, 0, 0)
  )
```

#### Publication Detail

The table below lists all retrieved articles. Use the search box to filter by keyword, journal, or date, and click column headers to sort.

```{r}
#| label: pubmed-article-table
#| eval: true
#| code-fold: true

library(DT)

articles |>
  mutate(
    # Create clickable PubMed link
    PMID = ifelse(
      !is.na(pmid),
      paste0('<a href="https://pubmed.ncbi.nlm.nih.gov/', pmid,
             '/" target="_blank">', pmid, '</a>'),
      pmid
    ),
    # Create clickable DOI link
    DOI = ifelse(
      !is.na(doi),
      paste0('<a href="https://doi.org/', doi,
             '" target="_blank">', doi, '</a>'),
      ""
    )
  ) |>
  select(
    PMID, Title = title, Journal = journal,
    `Publication Date` = pub_date, DOI
  ) |>
  datatable(
    caption = paste0(
      "PubMed: Opioid Overdose Fentanyl Research (",
      format(date_start, "%Y"), "-", format(date_end, "%Y"), ")"
    ),
    filter = "top",
    escape = FALSE,
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(3, "desc")),
      columnDefs = list(
        list(width = "400px", targets = 1)
      )
    ),
    rownames = FALSE
  )
```

::: {.callout-tip title="Reproducibility"}
For higher rate limits, register for a free NCBI API key at [ncbi.nlm.nih.gov/account](https://www.ncbi.nlm.nih.gov/account/) and append `&api_key=YOUR_KEY` to requests. The E-utilities support complex Boolean queries using MeSH terms (e.g., `"Opiate Overdose"[MeSH] AND "Fentanyl"[MeSH]`) for more precise results.
:::

### NIH RePORTER (Funded Research Projects)

[NIH RePORTER](https://reporter.nih.gov/) tracks all NIH-funded research projects, including active grants, publications, patents, and clinical trials. Its [API](https://api.reporter.nih.gov/) supports POST-based queries to identify what research is currently being funded on any topic.

```{r}
#| label: nih-reporter-search
#| eval: true
#| cache: true
#| cache.extra: !expr Sys.Date()
#| code-fold: true

# NIH RePORTER API: find funded opioid overdose research projects
# Documentation: https://api.reporter.nih.gov/

reporter_url <- "https://api.reporter.nih.gov/v2/projects/search"

# Paginate to collect all available projects
all_projects <- list()
for (pg in 0:99) {
  reporter_body <- list(
    criteria = list(
      advanced_text_search = list(
        operator = "and",
        search_field = "projecttitle,terms",
        search_text = "opioid overdose fentanyl"
      ),
      fiscal_years = as.list((year_end - 4L):year_end),
      exclude_subprojects = TRUE
    ),
    offset = pg * 100,
    limit = 100,
    sort_field = "FiscalYear",
    sort_order = "desc"
  )

  reporter_response <- request(reporter_url) |>
    req_body_json(reporter_body) |>
    req_perform()

  reporter_data <- reporter_response |>
    resp_body_json()

  if (length(reporter_data$results) == 0) break
  all_projects[[pg + 1]] <- reporter_data$results
  if (length(reporter_data$results) < 100) break
}

reporter_total <- reporter_data$meta$total

# Parse all results into a dataframe
projects <- lapply(unlist(all_projects, recursive = FALSE), function(p) {
  tibble(
    project_num   = p$project_num %||% NA_character_,
    fiscal_year   = p$fiscal_year %||% NA_integer_,
    title         = p$project_title %||% NA_character_,
    pi_name       = {
      if (!is.null(p$principal_investigators) && length(p$principal_investigators) > 0)
        p$principal_investigators[[1]]$full_name
      else NA_character_
    },
    organization  = p$organization$org_name %||% NA_character_,
    award_amount  = p$award_amount %||% NA_real_,
    abstract_text = p$abstract_text %||% NA_character_
  )
}) |>
  bind_rows()
```

The search matched **`r format(reporter_total, big.mark = ",")`** projects. **`r format(nrow(projects), big.mark = ",")`** were retrieved, representing **$`r format(sum(projects$award_amount, na.rm = TRUE), big.mark = ",")`** in total funding.

#### NIH Opioid Research Funding Over Time

The chart below shows the distribution of NIH-funded opioid research project awards by fiscal year. Hover over each point to see the project title, PI, and award amount.

```{r}
#| label: nih-reporter-timeline
#| eval: true
#| code-fold: true

library(plotly)

# Summarize funding by fiscal year
nih_annual <- projects |>
  filter(!is.na(award_amount), !is.na(fiscal_year)) |>
  group_by(fiscal_year) |>
  summarise(
    total_funding = sum(award_amount, na.rm = TRUE),
    n_projects = n(),
    .groups = "drop"
  ) |>
  arrange(fiscal_year)

nih_timeline_plot <- plot_ly() |>
  add_trace(
    data = nih_annual,
    x = ~fiscal_year,
    y = ~total_funding,
    type = "bar",
    marker = list(color = "#2494f7"),
    text = ~paste0(
      "<b>FY ", fiscal_year, "</b><br>",
      "Projects: ", n_projects, "<br>",
      "Total funding: $", formatC(total_funding, format = "f",
                                   digits = 0, big.mark = ",")
    ),
    hoverinfo = "text"
  ) |>
  layout(
    title = list(
      text = "NIH Opioid Overdose Research Funding by Fiscal Year",
      font = list(size = 16)
    ),
    xaxis = list(
      title = "Fiscal Year",
      dtick = 1
    ),
    yaxis = list(
      title = "Total Award Amount ($)",
      separatethousands = TRUE
    ),
    hovermode = "x unified"
  ) |>
  config(displayModeBar = TRUE)

nih_timeline_plot
```

#### What Is NIH Funding?

The word cloud below visualizes the most frequent terms appearing in project abstracts across all `r nrow(projects)` retrieved NIH-funded opioid research projects, revealing the dominant research themes.

```{r}
#| label: nih-reporter-wordcloud
#| eval: true
#| code-fold: true
#| column: page
#| fig-width: 14
#| fig-height: 4
#| out-width: "100%"

# Tokenize abstracts into individual words, remove stop words and generic terms
nih_word_counts <- tibble(text = projects$abstract_text) |>
  filter(!is.na(text), text != "") |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(
    !word %in% c("study", "research", "project", "aim", "aims",
                  "specific", "propose", "proposed", "will", "use",
                  "data", "based", "including", "provide", "develop",
                  "support", "national", "health", "program", "grant",
                  "funding", "abstract", "1", "2", "3", "4", "5",
                  "r01", "r21", "r34", "k01", "k23", "u01", "p50"),
    nchar(word) > 2,
    !grepl("^[0-9]+$", word)
  ) |>
  count(word, sort = TRUE)

ggplot(nih_word_counts |> head(80), aes(label = word, size = n, color = n)) +
  geom_text_wordcloud_area(family = "sans", shape = "square") +
  scale_size_area(max_size = 28) +
  scale_color_gradient(low = "#204d70", high = "#2494f7") +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(0, 0, 0, 0)
  )
```

#### Project Detail

The table below lists all retrieved NIH-funded projects. Use the search box to filter by keyword, PI, organization, or fiscal year, and click column headers to sort.

```{r}
#| label: nih-reporter-table
#| eval: true
#| code-fold: true

library(DT)

projects |>
  mutate(
    # Create clickable link to NIH RePORTER project page
    `Project Number` = ifelse(
      !is.na(project_num),
      paste0('<a href="https://reporter.nih.gov/project-details/',
             gsub(" ", "", project_num),
             '" target="_blank">', project_num, '</a>'),
      project_num
    ),
    `Award Amount` = award_amount,
    # Truncate abstracts for display
    Abstract = ifelse(
      !is.na(abstract_text) & nchar(abstract_text) > 200,
      paste0(substr(abstract_text, 1, 200), "..."),
      abstract_text
    )
  ) |>
  select(
    `Project Number`, `Fiscal Year` = fiscal_year,
    Title = title, PI = pi_name,
    Organization = organization, `Award Amount`, Abstract
  ) |>
  datatable(
    caption = paste0(
      "NIH-Funded Opioid Overdose Research Projects (",
      year_end - 4, "-", year_end, ")"
    ),
    filter = "top",
    escape = FALSE,
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(5, "desc")),
      columnDefs = list(
        list(width = "300px", targets = 2),
        list(width = "300px", targets = 6)
      )
    ),
    rownames = FALSE
  ) |>
  formatCurrency("Award Amount", digits = 0)
```

::: {.callout-note title="Data Source"}
NIH RePORTER is the authoritative source for federally funded biomedical research. It covers grants from NIH, CDC, FDA, AHRQ, SAMHSA, and other HHS agencies. No authentication is required. The API returns project details, funding amounts, and links to resulting publications.
:::

### Europe PMC (Open Access Literature)

[Europe PMC](https://europepmc.org/) provides open access to a worldwide collection of life science publications, including all PubMed content plus preprints, patents, and clinical guidelines. Its [REST API](https://europepmc.org/RestfulWebService) is particularly useful for retrieving full-text open access articles.

```{r}
#| label: europe-pmc-search
#| eval: true
#| cache: true
#| cache.extra: !expr Sys.Date()
#| code-fold: true

# Paginate to collect all available results (pageSize max is 1000)

epmc_url <- "https://www.ebi.ac.uk/europepmc/webservices/rest/search"

epmc_all <- list()
for (pg in 1:50) {
  epmc_response <- request(epmc_url) |>
    req_url_query(
      query      = paste0("(opioid overdose fentanyl) AND (PUB_YEAR:[",
                          year_start, " TO ", year_end, "])"),
      format     = "json",
      pageSize   = 1000,
      cursorMark = if (pg == 1) "*" else epmc_cursor,
      resultType = "core"
    ) |>
    req_perform()

  epmc_data <- epmc_response |>
    resp_body_json()

  if (length(epmc_data$resultList$result) == 0) break
  epmc_all[[pg]] <- epmc_data$resultList$result
  epmc_cursor <- epmc_data$nextCursorMark %||% NULL
  if (is.null(epmc_cursor) || length(epmc_data$resultList$result) < 1000) break
}

epmc_hit_count <- epmc_data$hitCount

epmc_articles <- lapply(unlist(epmc_all, recursive = FALSE), function(a) {
  tibble(
    pmid            = a$pmid %||% NA_character_,
    title           = a$title %||% NA_character_,
    journal         = a$journalTitle %||% NA_character_,
    year            = a$pubYear %||% NA_character_,
    first_pub_date  = a$firstPublicationDate %||% NA_character_,
    is_open_access  = a$isOpenAccess %||% "N",
    cited_by_count  = a$citedByCount %||% 0L,
    doi             = a$doi %||% NA_character_,
    abstract        = a$abstractText %||% NA_character_
  )
}) |>
  bind_rows()
```

The search matched **`r format(epmc_hit_count, big.mark = ",")`** results. **`r format(nrow(epmc_articles), big.mark = ",")`** articles were retrieved, of which **`r sum(epmc_articles$is_open_access == "Y", na.rm = TRUE)`** are open access.

#### Europe PMC: Publication Timeline

The chart below shows the volume of opioid overdose fentanyl research indexed in Europe PMC by publication year.

```{r}
#| label: europe-pmc-timeline
#| eval: true
#| code-fold: true

library(plotly)

epmc_by_year <- epmc_articles |>
  filter(!is.na(.data$year)) |>
  mutate(pub_year = as.integer(.data$year)) |>
  count(pub_year, name = "n_articles") |>
  arrange(pub_year)

epmc_timeline_plot <- plot_ly(
  data = epmc_by_year,
  x = ~pub_year,
  y = ~n_articles,
  type = "bar",
  marker = list(color = "#00a4bb"),
  text = ~paste0("<b>", pub_year, "</b><br>Articles: ", n_articles),
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = "Europe PMC: Opioid Overdose Fentanyl Publications by Year",
      font = list(size = 16)
    ),
    xaxis = list(title = "Publication Year", dtick = 1),
    yaxis = list(title = "Number of Articles"),
    hovermode = "x unified"
  ) |>
  config(displayModeBar = TRUE)

epmc_timeline_plot
```

#### Europe PMC: Research Themes

The word cloud below visualizes the most frequent terms in article titles and abstracts retrieved from Europe PMC.

```{r}
#| label: europe-pmc-wordcloud
#| eval: true
#| code-fold: true
#| column: page
#| fig-width: 14
#| fig-height: 4
#| out-width: "100%"

epmc_word_counts <- epmc_articles |>
  mutate(text = paste(title, abstract, sep = " ")) |>
  select(text) |>
  filter(!is.na(text), text != " ") |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(
    !word %in% c("study", "results", "background", "methods",
                  "conclusions", "objective", "purpose", "findings",
                  "data", "based", "including", "conclusion",
                  "aim", "abstract", "introduction"),
    nchar(word) > 2,
    !grepl("^[0-9]+$", word)
  ) |>
  count(word, sort = TRUE)

ggplot(epmc_word_counts |> head(80), aes(label = word, size = n, color = n)) +
  geom_text_wordcloud_area(family = "sans", shape = "square") +
  scale_size_area(max_size = 28) +
  scale_color_gradient(low = "#204d70", high = "#2494f7") +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(0, 0, 0, 0)
  )
```

#### Europe PMC: Article Detail

The table below lists all retrieved articles. Use the search box to filter by keyword, journal, year, or open access status.

```{r}
#| label: europe-pmc-table
#| eval: true
#| code-fold: true

library(DT)

epmc_articles |>
  mutate(
    PMID = ifelse(
      !is.na(pmid),
      paste0('<a href="https://europepmc.org/article/med/', pmid,
             '" target="_blank">', pmid, '</a>'),
      ""
    ),
    DOI = ifelse(
      !is.na(doi),
      paste0('<a href="https://doi.org/', doi,
             '" target="_blank">', doi, '</a>'),
      ""
    ),
    `Open Access` = ifelse(is_open_access == "Y", "Yes", "No")
  ) |>
  select(
    PMID, Title = title, Journal = journal,
    Year = year, `Open Access`, `Cited By` = cited_by_count, DOI
  ) |>
  datatable(
    caption = paste0(
      "Europe PMC: Opioid Overdose Fentanyl Research (",
      year_start, "-", year_end, ")"
    ),
    filter = "top",
    escape = FALSE,
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(5, "desc")),
      columnDefs = list(
        list(width = "400px", targets = 1)
      )
    ),
    rownames = FALSE
  )
```

### ClinicalTrials.gov (Clinical Research)

[ClinicalTrials.gov](https://clinicaltrials.gov/) registers clinical studies on a wide range of conditions, including opioid use disorder treatment trials, naloxone distribution studies, and harm reduction interventions. The [v2 API](https://clinicaltrials.gov/data-api/about-api) supports structured queries.

```{r}
#| label: clinicaltrials-search
#| eval: true
#| cache: true
#| cache.extra: !expr Sys.Date()
#| code-fold: true

# ClinicalTrials.gov v2 API: find opioid-related clinical trials
# Documentation: https://clinicaltrials.gov/data-api/about-api
# Paginate using nextPageToken to collect up to ~200 trials

ct_url <- "https://clinicaltrials.gov/api/v2/studies"

all_trials_raw <- list()
next_token <- NULL
for (pg in 1:10) {
  ct_query <- list(
    `query.cond`  = "opioid overdose",
    `query.intr`  = "fentanyl OR naloxone OR buprenorphine",
    `filter.overallStatus` = "RECRUITING,ACTIVE_NOT_RECRUITING,COMPLETED",
    countTotal    = "true",
    pageSize      = 100,
    sort          = "LastUpdatePostDate:desc"
  )
  if (!is.null(next_token)) ct_query$pageToken <- next_token

  ct_response <- request(ct_url) |>
    req_url_query(!!!ct_query) |>
    req_perform()

  ct_data <- ct_response |>
    resp_body_json()

  if (length(ct_data$studies) == 0) break
  all_trials_raw[[pg]] <- ct_data$studies
  next_token <- ct_data$nextPageToken
  if (is.null(next_token)) break
}

ct_total <- ct_data$totalCount

trials <- lapply(unlist(all_trials_raw, recursive = FALSE), function(s) {
  proto <- s$protocolSection
  id_mod <- proto$identificationModule
  status_mod <- proto$statusModule
  sponsor_mod <- proto$sponsorCollaboratorsModule
  design_mod <- proto$designModule
  desc_mod <- proto$descriptionModule

  tibble(
    nct_id      = id_mod$nctId %||% NA_character_,
    title       = id_mod$briefTitle %||% NA_character_,
    status      = status_mod$overallStatus %||% NA_character_,
    start_date  = status_mod$startDateStruct$date %||% NA_character_,
    sponsor     = sponsor_mod$leadSponsor$name %||% NA_character_,
    enrollment  = design_mod$enrollmentInfo$count %||% NA_integer_,
    description = desc_mod$briefSummary %||% NA_character_
  )
}) |>
  bind_rows()
```

The search matched **`r format(ct_total, big.mark = ",")`** studies. **`r nrow(trials)`** trials were retrieved for analysis.

#### Clinical Trials Over Time

The chart below shows the number of opioid-related clinical trials by their start year, broken down by status.

```{r}
#| label: clinicaltrials-timeline
#| eval: true
#| code-fold: true

library(plotly)

trials_timeline <- trials |>
  mutate(
    start_year = as.integer(substr(start_date, 1, 4))
  ) |>
  filter(!is.na(start_year)) |>
  count(start_year, status, name = "n_trials") |>
  arrange(start_year)

# Brand-aligned status colors
status_colors <- c(
  "COMPLETED" = "#2494f7",
  "ACTIVE_NOT_RECRUITING" = "#00a4bb",
  "RECRUITING" = "#204d70"
)

ct_timeline_plot <- plot_ly(
  data = trials_timeline,
  x = ~start_year,
  y = ~n_trials,
  color = ~status,
  colors = status_colors,
  type = "bar",
  text = ~paste0(
    "<b>", start_year, "</b><br>",
    status, ": ", n_trials, " trials"
  ),
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = "Opioid-Related Clinical Trials by Start Year and Status",
      font = list(size = 16)
    ),
    xaxis = list(title = "Start Year", dtick = 1),
    yaxis = list(title = "Number of Trials"),
    barmode = "stack",
    legend = list(
      title = list(text = "Trial Status"),
      orientation = "h",
      y = -0.2, x = 0
    ),
    hovermode = "x unified"
  ) |>
  config(displayModeBar = TRUE)

ct_timeline_plot
```

#### Clinical Trial Research Themes

The word cloud below visualizes the most frequent terms in trial titles and descriptions, revealing the dominant intervention and population themes in clinical research.

```{r}
#| label: clinicaltrials-wordcloud
#| eval: true
#| code-fold: true
#| column: page
#| fig-width: 14
#| fig-height: 4
#| out-width: "100%"

ct_word_counts <- trials |>
  mutate(text = paste(title, description, sep = " ")) |>
  select(text) |>
  filter(!is.na(text), text != " ") |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(
    !word %in% c("study", "trial", "clinical", "phase",
                  "participants", "purpose", "determine",
                  "evaluate", "assess", "including", "based",
                  "aims", "aim", "objective", "primary"),
    nchar(word) > 2,
    !grepl("^[0-9]+$", word)
  ) |>
  count(word, sort = TRUE)

ggplot(ct_word_counts |> head(80), aes(label = word, size = n, color = n)) +
  geom_text_wordcloud_area(family = "sans", shape = "square") +
  scale_size_area(max_size = 28) +
  scale_color_gradient(low = "#204d70", high = "#2494f7") +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(0, 0, 0, 0)
  )
```

#### Clinical Trial Detail

The table below lists all retrieved clinical trials. Use the filters to explore by status, sponsor, or keyword.

```{r}
#| label: clinicaltrials-table
#| eval: true
#| code-fold: true

library(DT)

trials |>
  mutate(
    `NCT ID` = ifelse(
      !is.na(nct_id),
      paste0('<a href="https://clinicaltrials.gov/study/',
             nct_id, '" target="_blank">', nct_id, '</a>'),
      nct_id
    ),
    # Truncate descriptions for display
    Description = ifelse(
      !is.na(description) & nchar(description) > 200,
      paste0(substr(description, 1, 200), "..."),
      description
    )
  ) |>
  select(
    `NCT ID`, Title = title, Status = status,
    `Start Date` = start_date, Sponsor = sponsor,
    Enrollment = enrollment, Description
  ) |>
  datatable(
    caption = "ClinicalTrials.gov: Opioid Overdose Intervention Trials",
    filter = "top",
    escape = FALSE,
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      order = list(list(3, "desc")),
      columnDefs = list(
        list(width = "300px", targets = 1),
        list(width = "300px", targets = 6)
      )
    ),
    rownames = FALSE
  )
```

::: {.callout-tip title="Reproducibility"}
ClinicalTrials.gov does not require authentication. The v2 API returns JSON by default and supports filtering by condition, intervention, status, location, and date range. This is useful for mapping the research landscape around specific interventions (e.g., naloxone distribution, buprenorphine access).
:::

# References